---
title: "Curricular Choke Points"
params:
  major_of_interest: CS
  n_persist: 8
  output: ~/ENGE/Studies/CSMigration/figures
  plot_height: 3
  plot_width: 6.5
  tikz_output: ~/Documents/manuscript2/figures
  svg_output: ~/ENGE/defense/img
  svg_width: 7
  svg_height: 4
output:
  html_document:
    df_print: paged
---

The focus of this report is the role the intro courses CS 1114 and CS 2114 play on students' matriculation into CS, leaving CS, and earning a CS degree.

```{r setup}
library(tidyverse)
library(tikzDevice)
library(survival)
library(survminer)
library(cmprsk)

library(vtir)
library(racadia)
library(vtirdata)
library(COEanalysis)
```


```{r load-data, cache=TRUE, results=FALSE}
vtir_terms %>%
  filter(student_level < 69, college_code == 5 | major == "CEM") ->
  coe_student_records

vtir_degrees %>% 
  filter_coe_degrees(coe_undergrad_data) ->
  coe_degree_data

vtir_terms %>%
  semi_join(coe_undergrad_records, by = "id") %>%
  filter(major == "CS") ->
  ever_in_cs

coe_undergrad_data %>%
  semi_join(vtir_courses %>%
              filter(subject == "CS", number == "1944"), by = "id") ->
  cs1_seminar

max_term <- max(vtir_terms$term, na.rm = TRUE)
```

## Entry Events

Entry events occure when a student enters a degree granting major. In the context of the COE and FYE, students may enter CS, another COE major, or leave the COE for a different major.
```{r cs entry events}
coe_undergrad_data %>%
  add_composit_vars(coe_undergrad_records) %>%
  left_join(cs1and2_spread, by = "id") %>%
  mutate(event = if_else(!is.na(matriculation_major), 
                         case_when(matriculation_major == "CS" ~ "MatCS",
                                   is_eng_major(matriculation_major) ~ "MatCOEMajor",
                                   TRUE ~ "LeaveCOE"),
                         if_else(term_enter >= max_term - 100,
                                 "InProgress",
                                 "LeaveCOE")),
         term = as_term(if_else(!is.na(matriculation_major), 
                        as.integer(matriculation_term), as.integer(last_term))),
         origin = fct_relevel(origin, "Traditional"),
         CS1_performance = grade_attempt_groups(CS1_grouped_grade, CS1_attempts, CS1_AP, CS1_transfer),
         CS2_performance = grade_attempt_groups(CS2_grouped_grade, CS2_attempts, CS2_AP, CS2_transfer)) %>%
    left_join(coe_undergrad_records %>% select(id, term, term_count), by = c("id", "term")) %>%
    mutate(periods = cut(term_count, 
                       breaks = c(0, 1.33, 2.66, 3.33, 3.66, 4.33, 4.66, 5.33, 5.66, 6.33, 6.66, 7.33, 7.66, 8.33, 24))) ->
  cs_entry_events

cs_entry_events %>%
  semi_join(cs1_seminar, by = "id") ->
  cs_intent_entry_events

xtabs(~ CS1_performance + event, cs_entry_events)
```

## Matriculation into CS

```{r}
library(survival)
library(survminer)
cox_fit_matcs <- coxph(Surv(as.numeric(periods), event == "MatCS") ~ gender + urm + first_generation, 
                       data = cs_entry_events)
coef(cox_fit_matcs)
p1 <- ggforest(cox_fit_matcs, data = cs_entry_events) +
  labs(title = "Hazards ratio for Matriculation into CS")
p1
```

```{r cox matriculate cs by course}
# library(eha)
# eha_fit <- coxreg(Surv(term_count, event == "MatCS") ~ gender + urm + first_generation, data = cs_entry_events %>% filter(event != "InProgress"))
# summary(eha_fit)
# exp(coef(eha_fit))

library(survival)
library(survminer)
cox_fit_matcs_cs1 <- coxph(Surv(term_count, event == "MatCS") ~ gender + urm + first_generation + 
                             CS1_AP + CS1_performance, data = cs_entry_events)
coef(cox_fit_matcs_cs1)
coxph_matcs_cs1_p <- ggforest(cox_fit_matcs_cs1, data = cs_entry_events) + labs(title = "Hazards ratio for Matriculation into CS")
coxph_matcs_cs1_p
```

```{r cox matriculate cs smaller}
# library(eha)
# eha_fit <- coxreg(Surv(term_count, event == "MatCS") ~ gender + urm + first_generation, data = cs_entry_events %>% filter(event != "InProgress"))
# summary(eha_fit)
# exp(coef(eha_fit))
cs_entry_events %>%
  mutate(CS1_performance = grade_attempt_groups(CS1_grouped_grade, CS1_attempts, CS1_AP, CS1_transfer)) ->
  cs_entry_events
cox_fit_matcs_cs1 <- coxph(Surv(term_count, event == "MatCS") ~ gender + urm + first_generation + 
                             CS1_performance, data = cs_entry_events)
                             
coxph_matcs_cs1_p <- ggforest(cox_fit_matcs_cs1, data = cs_entry_events) + labs(title = "Hazards ratio for Matriculation into CS")
ggsavetikz(coxph_matcs_cs1_p, file.path(params$tikz_output, "coxph_matcs_cs1.tex"), standAlone = TRUE, sanitize = TRUE, verbose = FALSE)
coxph_matcs_cs1_p
```

### Cumulative Incidents

```{r cumulative incidents setup}
library(cmprsk)
library(curry)
```

```{r cumulative incidents cs1 transfer}
gggrouped_cmprisk(cs_entry_events, CS1_performance, "cs_entry", list(
  title = "CS Entry Events and CS1 Seat",
  color = "CS1 Performance"
), saveTikz = TRUE)
```

Students who transfer in CS1 credit tend to leave the college at similar rates as everyone else after term 4, however before term 4 the have a lower probability of leaving the college. This could be due to the effect of experiencing CS1 as a time to evaluate one's ability to complete a CS degree


```{r cumulative incidents cs2 grade}
gggrouped_cmprisk(cs_entry_events, CS2_performance, "cs_entry", list(
  title = "CS Entry Events and CS2 Grade",
  color = "CS2 Performance"
))
```

## Exit Events

```{r cs exit events}
max_degree_term <- max(vtir_degrees$term_degree, na.rm = TRUE)
graduation_max_term <- max_degree_term- 600

coe_undergrad_data %>%
  inner_join(coe_undergrad_records %>%
               filter(major == "CS") %>%
               group_by(id) %>%
               dplyr::summarize(cs_matriculation = dplyr::first(term, order_by = term)) %>%
               ungroup(), by = "id") ->
  cs_students 

cs_students %>%
  select(id, cs_matriculation) %>%
  #has_degree("CS") %>%
  left_join(coe_degree_data %>% 
              filter(major == "CS") %>%
              select(id, term_degree, years_to_degree, major), by = "id") %>%
  mutate(hasDegree = !is.na(years_to_degree)) ->
  cs_degrees

# records for students who were ever in CS
coe_undergrad_records %>% 
  semi_join(cs_students, by = "id") %>% 
  left_join(cs_students %>% select(id, cs_matriculation), by = "id") -> 
  cs_records

cs_records %>%
  group_by(id) %>%
  dplyr::summarize(
    last_term = dplyr::last(term, order_by = term),
    last_major = dplyr::last(major, order_by = term)) ->
  cs_last_term

#eng_undergrad_records %>% filter(id == "18725") %>% select(id, term, student_level_desc, major) %>% head(20)
  
cs_records %>%
  anti_join(cs_degrees %>% filter(hasDegree), by = "id") %>%
  filter(term >= cs_matriculation, major != "CS") %>%
  group_by(id) %>%
  dplyr::summarize(exit_major = dplyr::first(major, order_by = term),
                   exit_term = dplyr::first(term, order_by = term)) ->
  leave_cs

coe_undergrad_records %>%
  semi_join(cs_students, by = "id") %>%
  group_by(id) %>%
  dplyr::summarize(last_term = dplyr::last(term, order_by = term),
                   last_major = dplyr::last(major, order_by = term)) ->
  last_term

leave_school <- last_term %>% 
  semi_join(cs_students, by = "id") %>%
  left_join(cs_students, by = "id") %>% 
  filter(as_term(last_term) < as_term(max(student_records$term))) %>% # student's who's last recorded term is less than the max we have for all students
  anti_join(cs_degrees %>% filter(hasDegree), by = "id") # and who have no CS degree

last_term %>% 
  semi_join(cs_students, by = "id") %>%
  filter(last_major == "CS") %>%
  anti_join(cs_degrees %>% filter(!hasDegree), by = "id") %>%
  anti_join(leave_school, by = "id") %>%
  left_join(coe_undergrad_data, by = "id") %>%
  filter(term_enter > graduation_max_term) -> 
  in_progress
  
rbind(cs_degrees %>% filter(!is.na(cs_matriculation), !is.na(term_degree)) %>% select(id, term = term_degree) %>% mutate(event = "CSDegree"),
      #cs_students %>% select(id, term = cs_matriculation) %>% mutate(event = "EnterCS"),
      leave_cs %>% select(id, term = exit_term) %>% mutate(event = "LeaveCS"),
      leave_school %>% select(id, term = last_term) %>% mutate(event = "LeaveCS"),
      in_progress %>% select(id, term = last_term) %>% mutate(event = "InProgress")) %>%
  distinct() ->
  cs_events

missing_records <- left_join(cs_degrees %>% 
                               filter(is.na(cs_matriculation)) %>% 
                               select(id, term_degree), coe_undergrad_data, by = "id")# %>% 
  #filter(term < degree_term, student_level <= 69)

cs_events %>%
  left_join(cs_students %>% 
              select(id, gender, urm, first_generation, term_enter), by = "id") %>%
  #left_join(cs_transfer_courses, by = "id") %>%
  mutate(cohort = term_year(term_enter)) %>%
  left_join(cs1and2_spread, by = "id") %>%
  left_join(coe_undergrad_records %>% select(id, term, term_count), by = c("id", "term")) %>%
  filter(gender != "Not Reported", !is.na(first_generation)) %>%
  droplevels() %>%
  filter(!is.na(term_count)) %>% 
  mutate(CS1_performance = grade_attempt_groups(CS1_grouped_grade, CS1_attempts, CS1_AP, CS1_transfer),
         CS2_performance = grade_attempt_groups(CS2_grouped_grade, CS2_attempts, CS2_AP, CS2_transfer))->
  cs_exit_events
```

## Cox Proportional Hazard Model for Exit Events

```{r cs exit cosph}
cox_fit_csdegree_demographics <- coxph(Surv(term_count, event == "CSDegree") ~ CS2_performance + urm + gender + first_generation, data = cs_exit_events)
coef(cox_fit_csdegree_demographics)
coxph_csdegree_demographics_p <- ggforest(cox_fit_csdegree_demographics, data = cs_exit_events) + 
  labs(title = "Hazards ratio for CS Degree")
ggsavetikz(coxph_csdegree_demographics_p, file.path(params$tikz_output, "coxph_csdegree_cs1.tex"), standAlone = TRUE, sanitize = TRUE, verbose = FALSE)
coxph_csdegree_demographics_p
```
```{r cs exit cosph cs1}
#cs_exit_events %>% mutate(CS1_seat = CS1_seat == "Yes") -> cs_exit_events
cox_fit_csdegree_cs1 <- coxph(Surv(term_count, event == "LeaveCS") ~ gender + urm + first_generation + CS1_performance, data = cs_exit_events)
coef(cox_fit_csdegree_cs1)
coxph_leavecs_cs1_p <- ggforest(cox_fit_csdegree_cs1, data = cs_exit_events) + 
  labs(title = "Hazards ratio for CS Degree")
ggsavetikz(coxph_leavecs_cs1_p, file.path(params$tikz_output, "coxph_leavecs_cs1.tex"), standAlone = TRUE, sanitize = TRUE, verbose = FALSE)
coxph_leavecs_cs1_p
```

```{r cs exit degree cosph cs2}
cox_fit_csdegree_cs2 <- coxph(Surv(term_count, event == "LeaveCS") ~ 
                                CS2_performance + gender + urm + first_generation, data = cs_exit_events
                              )
coef(cox_fit_csdegree_cs2)
coxph_csdegree_cs2_p <- ggforest(cox_fit_csdegree_cs2, data = cs_exit_events) + 
  labs(title = "Hazards ratio for CS Degree")
ggsavetikz(coxph_csdegree_cs2_p, file.path(params$tikz_output, "coxph_leavecs_cs2.tex"), standAlone = TRUE, sanitize = TRUE, verbose = FALSE)
coxph_csdegree_cs2_p
```

```{r cs exit leave cosph cs2}
cox_fit_leavecs_cs2 <- coxph(Surv(term_count, event == "LeaveCS") ~ 
                               gender + urm + first_generation + CS1_performance + CS2_performance, 
                             data = cs_exit_events  %>% droplevels())
coef(cox_fit_leavecs_cs2)
coxph_leavecs_cs2_p <- ggforest(cox_fit_leavecs_cs2, data = cs_exit_events) + 
  labs(title = "Hazards ratio for CS Degree")
ggsavetikz(coxph_leavecs_cs2_p, file.path(params$tikz_output, "coxph_leavecs_cs1and2.tex"), standAlone = TRUE, sanitize = TRUE, verbose = FALSE)
coxph_leavecs_cs2_p
```

```{r cs graduation cs1 attempts}
gggrouped_cmprisk(cs_exit_events, CS1_performance, "cs_exit", list(
  title = "CS Exit Events grouped by CS1 Performance",
  color = "CS1 Performance"
))
```

```{r cs graduation cs2 grade}
gggrouped_cmprisk(cs_exit_events, CS2_performance, "cs_exit", list(
  title = "CS Exit Events and CS2 Grade",
  color = "CS2 Grade Range"
))
```

```{r cs graduation cs1 transfer}
gggrouped_cmprisk(cs_exit_events, CS1_performance, "cs_exit", list(
  title = "CS Exit Events and CS1 Performance",
  color = "CS1 Seat"
), saveTikz = FALSE)
```

```{r cs graduation cs1 AP}
gggrouped_cmprisk(cs_exit_events, CS1_AP, "cs_exit", list(
  title = "CS Exit Events and CS AP Status",
  color = "CS AP"
), saveTikz = FALSE)
```

```{r cs graduation cs1 transfer}
gggrouped_cmprisk(cs_exit_events, cohort, "cs_exit", list(
  title = "CS Exit Events and CS Cohort",
  color = "Cohort"
), saveTikz = FALSE)
```
